Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/appendix.tex
editing

  This is obtained by composing the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
  with the square corresponding to the push property for $B$

@@

    Define a step-2 model $\mathcal M'$ as follows:
    \begin{itemize}
      \item Value objects are pairs consisting of:
      \begin{itemize}
        \item A value object $A$ in $\vf$.
        \item A commutative monoid $P_A$ of perturbations
        and a monoid homomorphism $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$.

@@

      \item Computation objects are pairs consisting of:
      \begin{itemize}
        \item A computation object $B$ in $\ef$.
        \item A commutative monoid $P_B$ of perturbations and a monoid homomorphism
        $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$.
      \end{itemize}

@@

      \item The objects of $\vsq'$ (i.e., the value relations) are pairs consisting of:
      \begin{itemize}
        \item A value relation $c \in \vsq$.
        \item A push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$.
        %\item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
      \end{itemize}

      The objects of $\esq'$ are defined analogously, i.e.,
      \begin{itemize}
        \item A computation relation $d \in \esq$.
        \item A push-pull structure $\pie_d$ for $d$ with respect to $P_B$ and $P_{B'}$.
        %\item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
      \end{itemize}

      \item The morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.

@@

    We define a ``Kleisli arrow'' operation taking a perturbation in $P_{UB}$
    to a perturbation in $P_{U(A \to B)}$, as follows:
    Given a perturbation $u = (n, b) \in P_{UB}$ we define $\id \tok u$ to be 
    $(n, \id_{P_A}, b) \in P_{U(A \to B)}$.
    We need to show that
%
    \[ \ptb_{U(A \arr B)}(\id \tok (n, b)) = \id \tok \ptb_{UB}(n, b). \]
%
    That is, we need to show that
%
    \[ (\delta_{U(A \arr B)}^*)^n \circ U(\id_{A} \arr \ptb_B(b)) = 
        \id \tok (\delta_{UB}^*)^n \circ U(\ptb_B(b)) \]
%
    This follows from the fact that $\delta_{UB}^*$ commutes with computation morphisms.
%

    Given a perturbation $q = (n, a) \in P_{FA}$ we define
    $(n, a) \tok \id = (n, a, \id_{P_B}) \in P_{U(A \to B)}$.
    We need to show that
  %
    \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
  %
    By similar reasoning to the above, this holds because $\delta_{FA}^*$ commutes with all computation morphisms.

@@

\begin{definition}[representation structure]
  Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure}
  $\rho^L_c$ for $c$ consists of a value morphism $e_c \in \vf(A, A')$ such that
  $c$ is quasi-left-representable by $e_c$ (see Definition \ref{def:quasi-left-representable}).
  We similarly define a {right-representation structure} $\rho^R_c$ for $c$ to consist of a morphism
  $p_c \in \vf{A', A}$ such that $c$ is quasi-right-representable by $p_c$ (see Definition \ref{def:quasi-right-representable}).
  % (Notice that the direction of the morphism is opposite in the definition of
  % right-representation structure.)

@@

    We define $\delrp_{F(c \comp c')}$ to be $\delrp_{Fc \comp Fc'} \circ \delta^r$.
    For $\dnl$, we paste the identity square $\delta^r \ltdyn \delta^r$ on top of
    the $\dnl$ square for the composition $Fc \comp Fc'$, and below that we paste
    the square $\id \ltdyn_{F(c \comp c')}^{Fc \comp Fc'} \id$ which we get from
    the lax property of $F$.

@@

  Likewise, let $\rho^R_{Fc_1}$ and $\rho^R_{Fc_2}$ be right-representation
  structures for $Fc_1$ and $Fc_2$ respectively.
  Then we can deinfe a right-representation structure for $F(c_1 \times c_2)$.
\end{lemma}
\begin{proof}
  We define $\rho^L_{c_1 \times c_2}$ as follows:
  \begin{itemize}
    \item $e_{c_1 \times c_2} = e_{c_1} \times e_{c_2}$
    \item $\delre_{c_1 \times c_2} = \delre_{c_1} \times \delre_{c_2}$ and
    likewise for $\delle_{c_1 \times c_2}$
    \item We get the commuting squares via the functorial action of $\times$
    on the corresponding squares for $c_1$ and $c_2$.
  \end{itemize}

@@

We define the functor $m$ which composes relations and squares horizontally as follows:
Let $c : A \rel A'$ and $c' : A' \rel A''$. We let

\begin{align*} 
  m ((c, \rho^L_c, \rho^R_{Fc}), (c', \rho^L_{c'}, \rho^R_{Fc'})) = \\
  (c \comp c', \rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}). 
\end{align*}

@@

\[ \Fsq^{\mathcal M'} (c, \rho^L_c, \rho^R_{Fc}) = 
  (\Fsq^M c, \rho^R_{Fc}, \rho^L_{UF(c)}), \]
and
\[ \Usq^{\mathcal M'} (d, \rho^R_d, \rho^L_{Ud}) = 
  (\Usq^M d, \rho^L_{Ud}, \rho^R_{FU(d)}), \]

@@

>>>>

  This is obtained by pasting $n$ copies of the square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$
  on top of the square corresponding to the push property for $c$.

@@

    Define a step-2 model $\mathcal M'$ as follows:
    \begin{itemize}
      \item Value objects are pairs consisting of:
      \begin{itemize}
        \item A value object $A$ in $\vf$.
        \item A monoid $P_A$ of perturbations
        and a monoid homomorphism $\ptb_A : P_A \to \{ f \in \vf(A, A) \mid f \bisim \id_A \}$.

@@

      \item Computation objects are pairs consisting of:
      \begin{itemize}
        \item A computation object $B$ in $\ef$.
        \item A monoid $P_B$ of perturbations and a monoid homomorphism
        $\ptb_B : P_B \to \{ \phi \in \ef(B, B) \mid \phi \bisim \id_B \}$.
      \end{itemize}

@@

      \item Given objects $(A, P_A, \ptb_A)$ and $(A', P_{A'}, \ptb_{A'})$ a value relation
      is a pair consisting of
      \begin{itemize}
        \item A value relation $c \in \vsq$.
        \item A push-pull structure $\piv_c$ for $c$ with respect to $P_A$ and $P_{A'}$.
        %\item A push-pull structure $\pie_{Fc}$ for $Fc$ with respect to $P^K_A$ and $P^K_{A'}$.
      \end{itemize}

      \item Similarly, a computation relation between $(B, P_B, \ptb_B)$ and $(B', P_{B'}, \ptb_{B'})$
      consists of
      \begin{itemize}
        \item A computation relation $d \in \esq$.
        \item A push-pull structure $\pie_d$ for $d$ with respect to $P_B$ and $P_{B'}$.
        %\item A push-pull structure $\piv_{Ud}$ for $Ud$ with respect to $P^K_B$ and $P^K_{B'}$.
      \end{itemize}

      \item The squares are the same as the squares of the original model $\mathcal M$
      %morphisms of $\vsq'$ and $\esq'$ are given by the morphisms of $\vsq$ and $\esq$.

@@

    We define a Kleisli arrow operation on perturbations that takes a perturbation
    in $P_{UB}$ to a perturbation in $P_{U(A \to B)}$, as follows:
    Given a perturbation $u = (n, b) \in P_{UB}$ we define $\id \tok u$ to be 
    $(n, \id_{P_A}, b) \in P_{U(A \to B)}$.
    We need to show that
%
    \[ \ptb_{U(A \arr B)}(\id \tok (n, b)) = \id \tok \ptb_{UB}(n, b). \]
%
    That is, we need to show that
%
    \[ (\delta_{U(A \arr B)}^*)^n \circ U(\id_{A} \arr \ptb_B(b)) = 
        \id \tok (\delta_{UB}^*)^n \circ U(\ptb_B(b)) \]
%
    This can be checked by unfolding the definition of $\tok$; the proof uses the fact that
    $\delta_{UB}^*$ commutes with computation morphisms.
%

    Given a perturbation $q = (n, a) \in P_{FA}$ we define
    $(n, a) \tok \id = (n, a, \id_{P_B}) \in P_{U(A \to B)}$.
    We need to show that
  %
    \[ \ptb_{U(A \arr B)}((n, a) \tok \id) = \ptb_{FA}(n, a) \tok \id. \]
  %
    By similar reasoning to the above, this holds because $\delta_{FA}^*$
    commutes with computation morphisms.

@@

\begin{definition}[representation structure]\label{def:representation-structure}
  Let $c : A \rel A'$ be a value relation. A \emph{left-representation structure}
  $\rho^L_c$ for $c$ consists of a value morphism $e_c \in \vf(A, A')$ such that
  $c$ is quasi-left-representable by $e_c$ (see Definition \ref{def:quasi-left-representable}).
  We similarly define a {right-representation structure} $\rho^R_c$ for $c$ to consist of a morphism
  $p_c \in \vf(A', A)$ such that $c$ is quasi-right-representable by $p_c$ (see Definition \ref{def:quasi-right-representable}).
  % (Notice that the direction of the morphism is opposite in the definition of
  % right-representation structure.)

@@

    We define $\delrp_{F(c \comp c')}$ to be $\delrp_{Fc \comp Fc'} \circ \delta^r$.
    For $\dnl$, we paste the identity square $\delta^r \ltdyn \delta^r$ on top of
    the $\dnl$ square for the composition $Fc \comp Fc'$, and below that we paste
    the square $\id \ltdyn_{F(c \comp c')}^{Fc \comp Fc'} \id$ which we get from
    the fact that $F$ is lax.

@@

  Likewise, let $\rho^R_{Fc_1}$ and $\rho^R_{Fc_2}$ be right-representation
  structures for $Fc_1$ and $Fc_2$ respectively.
  Then we can define a right-representation structure for $F(c_1 \times c_2)$.
\end{lemma}
\begin{proof}
  We define $\rho^L_{c_1 \times c_2}$ as follows:
  \begin{itemize}
    \item $e_{c_1 \times c_2} = e_{c_1} \times e_{c_2}$
    \item $\delre_{c_1 \times c_2} = \delre_{c_1} \times \delre_{c_2}$ and
    likewise for $\delle_{c_1 \times c_2}$
    \item We get the commuting squares via the functorial action of $\times$
    on the corresponding squares for $c_1$ and $c_2$.
  \end{itemize}

@@

We define horizontal composition of relations and squares as follows:
Let $c : A \rel A'$ and $c' : A' \rel A''$. We define
%
\begin{align*} 
  ((c, \rho^L_c, &\rho^R_{Fc}) \comp (c', \rho^L_{c'}, \rho^R_{Fc'})) =
  (c \comp c', \rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}), 
\end{align*}
%
and
%
\begin{align*}
  ((d, \rho^R_d, &\rho^L_{Ud}) \comp (d', \rho^R_{d'}, \rho^L_{Ud'})) =
  (d \comp d', \rho^R_{d \comp d'}, \rho^L_{U(d \comp d')}),
\end{align*}
%
where $\rho^L_{c \comp c'}, \rho^R_{F(c \comp c')}, \rho^R_{d \comp d'},$
and $\rho^L_{U(d \comp d')}$ are as defined in Lemma \ref{lem:representation-comp}.

@@

\[ F(c, \rho^L_c, \rho^R_{Fc}) = 
  (F c, \rho^R_{Fc}, \rho^L_{UF(c)}), \]
and
\[ U(d, \rho^R_d, \rho^L_{Ud}) = 
  (U d, \rho^L_{Ud}, \rho^R_{FU(d)}), \]

@@

We now establish the quasi-order-equivalence for the functors.
We already showed that $U(d \comp d') \qordeq U(d)U(d')$ and $F(c \comp c'') \qordeq F(c)F(c')$
in the proof of Lemma \ref{lem:representation-comp}.
The other two properties
$(cc') \to (dd') \qordeq (c \to d)(c' \to d')$ and
$(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
are proved similarly, noting in both cases that the both relations are
quasi-represented by the same morphism.

### 2024-01-27 paper-new/categorical-model.tex
work on proof of adequacy and move it to concrete model; work on dyn in concrete model
editing

This mathematical structure is sufficient to interpret the theory of
gradual typing in Section~\ref{sec:gtlc}.
\begin{definition}
  Given any extensional gradual typing model, we interpret
  \begin{enumerate}
  \item Each type $A$ as a value type, interpreting the base types and
    $\times$ as their semantic analogues and $\sem {A \ra A'}$ as $U(\sem{A} \to
    F(\sem{A'}))$.
  \item Every type precision derivation $c : A \ltdyn A'$ is
    interpreted as a relation morphism $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
    way. Every equivalence axiom $c \equiv c'$ implies that $\sem{c} \equidyn \sem{c'}$.
  \item Every term $\Gamma \vdash M : A$ is interpreted as a morphism
    $\sem{M} : \mathcal V_f (\times\sem{\Gamma},UF\sem{A})$. Upcasts are interpreted as
    $UF(u_{\sem{c}})$ and downcasts as $Ud_{F\sem{c}}$.
  \item If $\Delta \vdash M \ltdyn M' : c$ then $\sem{M}
    \ltsq{\times\sem{\Delta}}{\sem{c}} \sem{M'}$ holds.
  \end{enumerate}
\end{definition}

@@

We can specify this elegantly as a category internal to the category of CBPV models
and lax morphisms, where we require that the reflexivity, source, and target morphisms
are strict.
In particular, as in the extensional case, there is a CBPV model of ``objects''
$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
There are CBPV morphisms
$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
just as before.
%
But now, we also have a CBPV morphism $m$ from the pullback 
$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
In particular, this consists of a functor 
$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
for composition of computation relations/squares.
Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.

@@

For the sake of ease of reference, we recap the definition of a step-0 model:

\begin{definition}
  A \emph{step-0} model of intensional gradual typing consists of:
  \begin{itemize}
    \item A category internal to the category of CBPV models and lax morphisms,
    where we require that the morphisms $r$, $s$, and $t$ are strict.
    \item A natural transformation $\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$.
    \item A value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \end{itemize}

@@

% TODO explain what this means in terms of bisimilarity "squares"?
In this setting, we write $\refl_A : A \bisim A$ to mean that $\refl_A \in \ob(\vsim)$,
such that $\svsim(\refl_A) = A = \tvsim(\refl_A)$.
Let $f, f' \in \vf(A_i, A_o)$.
The judgment $\gamma : f \bisim_{A_i, A_o} f'$ is defined to mean:

@@

Lastly, we require that for any computation object $B$, the hom-set $\vf(UB, UB)$ contains a
distinguished morphism $\delta_{UB}^*$, such that $\delta_{UB}^* \bisim_{UB, UB} \id_{UB}$.
Moreover, we require that these morphisms are related in that for any
$d : B \rel B'$, we have a square $\delta_{UB}^* \ltdyn_{Ud}^{Ud} \delta_{UB'}^*$.
We also require that these morphisms commute with computation morphisms, in the sense
that for any $\phi \in \ef(B, B')$ we have $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$.

Given the existence of the morphisms $\delta_{UB}^*$, we can define a computation morphism
$\delta_{FA}^* \in \ef(FA, FA)$ for all $A$ by composing the unit $\eta_A \in \vf(A, UFA)$ with
the morphism $\delta_{UFA}^* \in \vf(UFA, UFA)$, and then by the adjunction
we get a computation morphism $\delta_{FA}^ \in \ef(FA, FA)$. Moreover, we get a square
$\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A$.

@@

We will slightly abuse notation and refer to an endomorphism $f \in \vf(A, A)$ as being ``in''
the monoid of perturbations, but which we actually mean there is an element $p \in P_A$
that is mapped to $f$ under the homomorphism.

We require that $\delta_{UB}^* \in P_{UB}$ for all $A$, where $\delta_{UB}^*$ is the distinguished
morphism that is required to be present in every hom-set $\vf(UB, UB)$ per the definition
of a step-1 model.

@@

We claim that from a step-1 model, we can construct a step-2 model. 
\eric{TODO give overview}
For the proof, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.

@@

Now suppose we have a step-2 intensional model $\mathcal{M}$.
We claim that we can construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.

\eric{TODO give overview}

For the proof, see Lemma \ref{lem:step-2-model-to-step-3-model} in the Appendix.

@@

Using our existing notation, we say that $f \le_{c_o}^{c_i} g$ if there exist $f'$ and $g'$ such that

>>>>

This mathematical structure is sufficient to interpret the theory of
gradual typing in Section~\ref{sec:GTLC}.
\begin{definition}
  Given any extensional gradual typing model, we interpret
  \begin{enumerate}
  \item Each type $A$ as a value type, interpreting the base types and
    $\times$ as their semantic analogues and $\sem {A \ra A'}$ as $U(\sem{A} \to
    F(\sem{A'}))$.
  \item Every type precision derivation $c : A \ltdyn A'$ is
    interpreted as a relation morphism $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
    way. Every equivalence axiom $c \equiv c'$ implies that $\sem{c} \equidyn \sem{c'}$.
  \item Every term $\Gamma \vdash M : A$ is interpreted as a morphism
    $\sem{M} : \mathcal V_f (\times\sem{\Gamma},UF\sem{A})$. Upcasts are interpreted as
    $UF(u_{\sem{c}})$ and downcasts as $Ud_{F\sem{c}}$.
  \item If $\Delta \vdash M \ltdyn M' : c$ then $\sem{M}
    \ltsq{\times\sem{\Delta}}{\sem{c}} \sem{M'}$ holds.
  \end{enumerate}
\end{definition}

@@

We can specify this compactly as a category internal to the category of CBPV models
and lax morphisms, where we require that the reflexivity, source, and target morphisms
are strict.
In particular, as in the extensional case, there is a CBPV model of ``objects''
$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
There are CBPV morphisms
$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
just as before.
%
But now, we also have a CBPV morphism $m$ from the pullback 
$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
In particular, this consists of a functor 
$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
for composition of computation relations/squares.
Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.

@@

For the sake of ease of reference, we recap the definition of a step-0 model:
%
\begin{definition}
  A \emph{step-0} model of intensional gradual typing consists of:
  \begin{itemize}
    \item A category internal to the category of CBPV models and lax morphisms,
    where we require that the morphisms $r$, $s$, and $t$ are strict.
    \item A natural transformation $\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$.
    \item A value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \end{itemize}

@@

In this setting, we write $\refl_A : A \bisim A$ to mean that $\refl_A \in \ob(\vsim)$,
such that $\svsim(\refl_A) = A = \tvsim(\refl_A)$.
Let $f, f' \in \vf(A_i, A_o)$.
The judgment $\gamma : f \bisim_{A_i, A_o} f'$ is defined to mean:

@@

Lastly, we require that for any computation object $B$, the hom-set $\vf(UB, UB)$ contains a
distinguished morphism $\delta_{UB}^*$, such that $\delta_{UB}^* \bisim_{UB, UB} \id_{UB}$.
Moreover, we require that these morphisms are related in that for any
$d : B \rel B'$, we have a square $\delta_{UB}^* \ltdyn_{Ud}^{Ud} \delta_{UB'}^*$.
We also require that these morphisms commute with computation morphisms, in the sense
that for any $\phi \in \ef(B, B')$ we have $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$.
%
Given the existence of the morphisms $\delta_{UB}^*$, we can define a computation morphism
$\delta_{FA}^* \in \ef(FA, FA)$ for all $A$ by composing the unit $\eta_A \in \vf(A, UFA)$ with
the morphism $\delta_{UFA}^* \in \vf(UFA, UFA)$, and then by the adjunction
we get a computation morphism $\delta_{FA}^ \in \ef(FA, FA)$. Moreover, we get a square
$\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A$.

@@

We will slightly abuse notation and refer to an endomorphism $f \in \vf(A, A)$ as being ``in''
the monoid of perturbations, by which we actually mean there is an element $p \in P_A$
that is mapped to $f$ under the homomorphism.

We require that $\delta_{UB}^* \in P_{UB}$ for all $B$, where $\delta_{UB}^*$ is the distinguished
morphism that is required to be present in every hom-set $\vf(UB, UB)$ per the definition
of a step-1 model.

@@

We claim that from a step-1 model, we can construct a step-2 model. 
The value objects of the model are defined to be triples $(A, P_A, \ptb_A)$ where $A$ is a
value object in $\mathcal{M}$, $P_A$ is a monoid and $\ptb_A$ is a homomorphism of monoids
from $P_A$ to the endomorphisms on $A$ that are bisimilar to the identity.
Likewise, computation objects are triples $(B, P_B, \ptb_B)$.
The morphisms are the same as the morphisms of $\mathcal{M}$.
A value relation between $(A, P_A, \ptb_A)$ and $(A', P_{A'}, \ptb_{A'})$ is given by
a pair of a relation in $c$ and a \emph{push-pull structure} $\Pi_c$ specifying that
that $c$ satisfies the push-pull property. Computation relations are defined analogously.
The squares are the same as those of $\mathcal{M}$.
We define the action of the functor $F$ on objects as $F(A, P_A, \ptb_A) = 
(FA, \mathbb{N} \times P_A, \ptb_{FA})$ where $\ptb_{FA}(n, a) = (\delta_{FA}^*)^n \circ F(\ptb_A(a))$
(i.e., we use the distinguished delay morphism $\delta_{FA}^*$).
The action of $U$ on objects is defined similarly, where the perturbations
are defined to be $\mathbb{N} \times P_B$.

For the full details of the construction, see Lemma \ref{lem:step-1-model-to-step-2-model} in the Appendix.

@@

Now suppose we have a step-2 intensional model $\mathcal{M}$.
We claim that we can construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
In the construction, the objects and morphisms are the same as those of $\mathcal M$,
and a value relation between $A$ and $A'$ consists of a triple $(c, \rho^L_c, \rho^R_{Fc})$
where $c : A \rel A'$ is a relation in $\mathcal{M}$,
$\rho^L_c$ is a quasi-\emph{left}-representation for $c$ 
(i.e., an embedding $e_c$, perturbations $\delre_c \in P_{A'}$ and $\delle_c \in P_A$, 
and the two relevant squares), and similarly $\rho^R_{Fc}$ is
a quasi-\emph{right}-representation for $Fc$.
Computation relations are triples $(d, \rho^R_d, \rho^L_{Ud})$ where $d : B \rel B'$,
$\rho^R_d$ is a quasi-right-representation for $d$ and $\rho^L{Ud}$ is a quasi-left-representation for $Ud$.
The value and computation squares are the same as those of $\mathcal{M}$.
We then define composition of relations and the action of the functors $F$, $U$, $\times$, and $\arr$.

For the full details of the construction, see Lemma \ref{lem:step-2-model-to-step-3-model} in the Appendix.

@@

Using our existing notation, we say that $f \ltls_{c_o}^{c_i} g$ if there exist $f'$ and $g'$ such that
